Linearity and tropical mathematics in the Lambda-calculus

### 2023-07-21 CSL24/section2.tex
Committ
Per ora è ancora molto copia ed incolla da LICS. Fai tutte le modifiche che vuoi, se vuoi. Io domani non posso lavorare, quindi prendi il lock quando vuoi. Da domenica, dimmelo così ci organizziamo. Notte !

In this section, we discuss in some more detail the two approaches to quantitative semantics we mentioned in the Introduction, at the same time providing an overview of how we aim at bridging them using tropical mathematics.

\subsection{Metric approach: bounded $\lambda$-terms as Lipschitz functions}

@@

In all these cases, it is common to consider forms of denotational semantics in which types are endowed with a \emph{behavioral metric}, that is, a metric on programs which account for differences in behavior. 
A fundamental insight coming from this line of work is that \emph{affine} programs, i.e.~programs that may use their input at most once, correspond to \emph{non-expansive} (or $1$-Lipschitz) maps, that is, to functions $f$ for which the distances
$d(f(x),f(y))$ produced in output are bounded by the distances $d(x,y)$ in input. 
A more formal way of stating this observation is that the category $\Met$ of pseudo-metric spaces and non-expansive maps provides a model of the \emph{linear} simply typed $\lambda$-calculus, being a \emph{symmetric monoidal closed} category, and in fact it also models {affine} terms (since the cartesian and monoidal units coincide).

@@

These observations have led to the study of linear $\lambda$-calculi with \emph{graded} exponentials, like $\mathsf{Fuzz}$ \cite{Reed2010}, inspired from Girard's Bounded Linear Logic \cite{Girard92tcs}, which have been applied to the study of differential privacy \cite{Gaboardi2013, Gaboardi2017}. The types of such systems are defined by combining linear constructors with a \emph{graded linear exponential comonad} $!_{r}(-)$ \cite{Katsumata2018}.
In the following sections we will sometimes make reference to a basic graded type system, that we call $\BSTLC$, for bounded higher-order programs, with types defined via $A::= o  \mid   !_{n}A\multimap A  $, 
where $n\in \BB N$. Intuitively, $!_{ n}A\multimap B$ is the type of functions from $A$ to $B$ that may use their input \emph{at most} $n$ times. More details about $\BSTLC$ are provided in the Appendix.

Now, what about the good old, ``unbounded'', simply typed $\lambda$-calculus? Actually, by using unbounded duplications, one might lose the Lipschitz property. For instance, while the functions $M_{k}=\lambda x. k\cdot x: \BB R\to \BB R$ are all Lipschitz-continuous, with Lipschitz constant $k$, the function $M=\lambda x.x^{2}$ obtained by ``duplicating'' $x$ is not Lipschitz anymore: $M$ is, so to say, \emph{too} sensitive to errors. 
More abstractly, it is well-known that the category $\Met$ is \emph{not} cartesian closed, so it is not a model of $\STLC$ (yet, several cartesian closed \emph{sub-}categories of $\Met$ do exist, see e.g.~\cite{Clementino2006, PistoneFSCD2022}).
Still, one might observe that the program $M$ above is actually Lipschitz-continuous, if not globally, at least \emph{locally} (i.e.~over any compact set). Indeed, some cartesian closed categories of locally Lipschitz maps have been produced in the literature \cite{Ehrhard2011, PistoneLICS}, and a new example will be exhibited in this paper.

\subsection{Resource approach: differential $\lambda$-terms as polynomials}

A different family of approaches to linearity and duplication arises from the study of the \emph{differential $\lambda$-calculus} \cite{difflambda} (and differential linear logic \cite{dill}) and its categorical models. 
The key ingredient is a \emph{differential operator} $\Der$,  added to the usual syntax of the $\lambda$-calculus. The intuition is that, given $M$ of type $A\to B$ and $N$ of type $A$, the program $\Der[M]\cdot N$, still of type $A\to B$, corresponds to the \emph{linear application} of $M$ to $N$: this means that $N$ is passed to $M$ so that the latter may use it exactly once (this is why $\Der[M]\cdot N$ still has type $A\to B$, since $M$ might need \emph{other} copies of an input of type $A$).

Interestingly, the categorical study of the operator $\Der$ has led to the introduction of \emph{cartesian differential categories} $C\partial C$ \cite{Blute2009}, a class of categories providing an abstract axiomatization of differentiation, in the usual mathematical sense. More precisely, a cartesian category $\C C$ is a $C\partial C$ when:
\begin{itemize}
\item $\C C$ is left-additive, i.e.~its hom-sets have the structure of commutative monoids, and the cartesian structure is well-behaved w.r.t.~this monoid structure;
\item $\C C$ is equipped with a differential operator $D:
\C C(X,Y)\to \C C(X\times X,Y)$ satisfying some axioms which capture usual properties of differentials (e.g.~the linearity of $D$ in one of its two variables, the chain rule, etc.).
\end{itemize}

Correspondingly, the syntax of the simply typed \emph{differential} $\lambda$-calculus ($\STDLC$) is defined by enriching $\STLC$ with a monoid structure $0,+$ over terms, as well as with $\Der$ and a notion of \emph{linear substitution} (see \cite{difflambda} or the Appendix for details).
The models of $\STDLC$ are the 
cartesian \emph{closed} differential categories ($CC\partial C$), 
which are defined as $C\partial C$ which are also cartesian closed, and in which the monoid structure and the differential operator are both well-behaved with respect to the closed structure \cite{Manzo2012}.

Another intriguing similarity between program derivatives and 
actual derivatives is provided by the \emph{Taylor expansion} $\C T$:
in $\STDLC$ one can expand any application $MN$ as an infinite formal sum of \emph{linear} applications
$\Der^{(k)}[M]\cdot N^{k}$, i.e.~where $N$ is passed exactly $k$ times to $M$:
\begin{align}\label{eq:taylor}
\Te{MN} :=  \sum_{k=0}^{\infty}\frac{1}{!k}\cdot (\Der^{(k)}[M] \cdot N^{k})0
\end{align}
In other words, unbounded duplications correspond to some sort of limit of bounded, but arbitrarily large, ones.% (this perspective is made clearer by the related approach of the \emph{resource $\lambda$-calculus} \cite{}).

@@

\subsection{Tropical Mathematics: a possible synthesis}%Outline of the paper}

At this point, as the Taylor formula decomposes an unbounded application as a limit of bounded ones, one might well ask whether it could be possible to see this formula as interpreting  a $\lambda$-term 
as a limit of Lipschitz maps, in some sense, thus bridging the metric and differential approaches.  
Here, a natural direction to look for is the \emph{relational semantics}, i.e.~the somehow canonical ``Taylor'' semantics for $\STDLC$. 
However, in this semantics, terms with bounded applications correspond to \emph{polynomials}, i.e.~to non-Lipschitz functions.

Yet, what if these polynomials were tropical ones, i.e.~piecewise linear functions? This way, \eqref{eq:taylor} could really be interpreted as a decomposition of $\lambda$-terms via limits (indeed, $\inf$s) of Lipschitz maps. In other words, unbounded term application could be seen 
as a limit of \emph{more and more sensitive} operations.

This viewpoint, that we develop in the following sections, leads to the somehow unexpected discovery of a bridge between the metric and differential study of higher-order programs.
This connection not only suggests the application of optimization methods based on tropical mathematics to the study of the $\lambda$-calculus and its quantitative extensions, but it scales to a 
more abstract level, leading to introduce a 
differential operator for continuous functors between \emph{generalized} metric spaces (in the sense of \cite{Lawvere1973}).

%
%At this point, 
%by interpreting such polynomials over the tropical semiring 
%
%Our question can thus be reformulated as follows: can we make the relational semantics \emph{Lipschitz}, hence amenable to metric and sensitivity analysis? The goal of this paper is to show that, by appealing to tropical mathematics, this is indeed possible and leads to the somehow unespected discovery of a bridge between the metric and differential study to higher-order programs.
%
%Fino a qui niente a capo
%
%Fino a qui niente a capo
%
%Fino a qui niente a capo
%
%Fino a qui niente a capo
%
%Fino a qui niente a capo
%
%Limite! Alla prossima riga sforo oltre 12 pag.
%

%In the next section we recall some basic ideas from tropical mathematics, and its connection with the study of the Lawvere quantale.
%Since polynomials correspond to piecewise linear (hence Lipschitz) functions in tropical mathematics,
%The reconstruction of the relational semantics over the tropical semi-ring, presented in Section \ref{section3} and Section \ref{section4}, will provide a metric semantics of the differential $\lambda$-calculus, bridging sensitivity and resource analysis. 
%In Section \ref{section5} we suggest potential applications of this approach, relating well-studied applications of program metrics, resource analysis with current uses of tropical mathematics in computer science.  
%Finally, in Section \ref{section6} we show that the connection between the metric and differential analysis of higher-order programs extends well beyond the relational semantics, through a more abstract correspondence between {generalized metric spaces} and modules over the tropical semiring.

>>>>

%In this section, we discuss in some more detail the two approaches to quantitative semantics we mentioned in the Introduction, at the same time providing an overview of how we aim at bridging them using tropical mathematics.

\subsection{Metric approach: bounded $\lambda$-calculus in a nutshell}

@@

%In all these cases, it is common to consider forms of denotational semantics in which types are endowed with a \emph{behavioral metric}, that is, a metric on programs which account for differences in behavior. 
A fundamental insight coming from this line of work is that \emph{affine} programs, i.e.~programs that may use their input at most once, correspond to \emph{non-expansive} (or $1$-Lipschitz) maps, that is, to functions $f$ for which the distances
$d(f(x),f(y))$ produced in output are bounded by the distances $d(x,y)$ in input. 
A more formal way of stating this observation is that the category $\Met$ of pseudo-metric spaces and non-expansive maps provides a model of the \emph{linear} simply typed $\lambda$-calculus, being a \emph{symmetric monoidal closed} category, and in fact it also models {affine} terms (since the cartesian and monoidal units coincide).

@@

These observations have led to the study of linear $\lambda$-calculi with \emph{graded} exponentials, like $\mathsf{Fuzz}$ \cite{Reed2010}, inspired from Girard's Bounded Linear Logic \cite{Girard92tcs}, which have been applied to the study of differential privacy \cite{Gaboardi2013, Gaboardi2017}. The types of such systems are defined by combining linear constructors with a \emph{graded linear exponential comonad} $!_{r}(-)$ \cite{Katsumata2018}.
The language $\BSTLC$ we consider is a simplified version of the language $\mathrm{Fuzz}$ \cite{Reed2010}. 
The terms are as for the $\STLC$, the types are $A::= o \ \mid  \ !_{n}A \multimap A$ and type judgements are of the form $\Gamma \vdash M:A$, where a context $\Gamma$ is a set of declarations of the form $x :_{n}A$, with $n\in \mathbb N$.
The typing rules are given in Fig.~\ref{fig:rules}, where $\Gamma+\Delta$ is defined letting $(\Gamma, x:_{m} A)+( \Delta, x:_{n} A) =  (\Gamma+\Delta), x:_{m+n}A$, and $m\Gamma$ is made all $x:_{mn}A$ for $(x:_{n}A) \in \Gamma$.  
The axiom is $x:_{1}A\vdash x: A$.
The main feature of this language is that if $\vdash \lambda x.M:\,!_nA\multimap B$, then $x$ will be duplicated exactly $n$ times in the reduction to the normal form. %, i.e.\ $\vdash \lambda x.M:\,!_nA\multimap B$ is a program which performs $n$ duplications.
E.g., $\vdash_{\BSTLC} \lambda {\color{red}z}.\left( \lambda x{\color{green}y}. yxx \right)z : \, !_{\color{red}2} o\multimap !_{\color{green}1}(!_{\color{violet}1} o \multimap !_{\color{blue}1} o \multimap o) \multimap o$.

The $\BSTLC$ can be interpreted in a symmetric monoidal closed category (SMCC) equipped with a \emph{$\mathbb N$-graded linear exponential comonad} \cite{Katsumata2018}.
%Observe that one can always type an affine term like e.g.~$\lambda xy.x$ with a linear type $A\multimap B\multimap A$. Instead, a term like $\lambda xy.x(xy)$ containing two occurrences of $x$ cannot be given the linear type $(A\multimap A)\multimap (A\multimap A)$ but a type of the form
%$!_{2}(A\multimap A)\multimap (A\multimap A)$.

\begin{figure*}

	\scriptsize

	\[ \arraycolsep=5pt\def\arraystretch{2.8}
	\begin{array}{cccc}
		\prooftree
		\Gamma \vdash M:A
		\justifies
		\Gamma, x:_{0}B \vdash M:A
		\endprooftree 
		&
		\prooftree
		\Gamma, x:_{n}B, y:_{m} B\vdash M:A
		\justifies
		\Gamma, x:_{n+m}B\vdash M[x/y]:A
		\endprooftree 
		&
		\prooftree
		\Gamma, x:_{n} A\vdash M: B
		\justifies
		\Gamma\vdash \lambda x.M: !_{n}A\multimap B
		\endprooftree
		&
		\prooftree
		\Gamma \vdash M: !_nA\multimap B
		\quad
		\Delta\vdash N: A
		\justifies
		\Gamma +n\Delta\vdash MN: B
		\endprooftree
		\\
		\\
		\hline
		\\
		\prooftree
		\Gamma, x: A\vdash M: B
		\justifies
		\Gamma\vdash \lambda x.M: A\to B
		\endprooftree 
		&
		\prooftree
		\Gamma \vdash M: A\to B
		\quad
		\Gamma\vdash \mathbb T: A
		\justifies
		\Gamma \vdash M\mathbb T: B
		\endprooftree 
		&
		\prooftree
		\Gamma \vdash M: A\to B
		\quad
		\Gamma \vdash N: A
		\justifies
		\Gamma \vdash \Diff{M}{N}: A\to B
		\endprooftree
		&
		\prooftree
		\Gamma\vdash M_1: A
		\,\cdots\,
		\Gamma \vdash M_n:A
		\justifies
		\Gamma \vdash M_1+\cdots +M_n : A
		\using (n\geq 2)
		\endprooftree
	\end{array}
	\]
	\caption{Typing rules (axiom rules are given in the text) for $\BSTLC$ (top) and $\STDLC$ (bottom).}\label{fig:rules}
\end{figure*}

\subsection{Resource approach: differential $\lambda$-calculus in a nutshell}

A different family of approaches to linearity and duplication arises from the study of the \emph{differential $\lambda$-calculus} \cite{difflambda} (and differential linear logic \cite{dill}) and its categorical models. 
The key ingredient is a \emph{differential operator} $\Der$,  added to the usual syntax of the $\lambda$-calculus. The intuition is that, given $M$ of type $A\to B$ and $N$ of type $A$, the program $\Diff{M}{N}$, still of type $A\to B$, corresponds to the \emph{linear application} of $M$ to $N$: this means that $N$ is passed to $M$ so that the latter may use it exactly once (this is why $\Diff{M}{N}$ still has type $A\to B$, since $M$ might need \emph{other} copies of an input of type $A$).

The syntax of the differential $\lambda$-calculus $\STDLC$ (e.g.\ \cite[Section 3]{Manzo2010}), is given by the \emph{terms} $M$ and the \emph{sums} $\mathbb T$, mutually generated by: $M::= x\mid \lambda x.M \mid M\mathbb T \mid \Diff{M}{M}$ and $\mathbb T::= 0 \mid M \mid M+\mathbb T$,
quotiented by: $\alpha$-equivalence; $+,0$ form a commutative monoid on the set of sums; %, i.e.\ commutativity and associativity of $+$ and neutrality of $0$ w.r.t.\ $+$;
linearity of $\lam x.(\_)$, $\Diff{\_}{\_}$ and $(\_)\mathbb T$ (but \emph{not} of $M(\_)$); irrelevance of the order of consecutive $\Diff{\_}{\_}$.
%Remark that $M(\_)$ is \emph{not} set to be linear: $\lambda x.0=0\mathbb T=\Diff{0}{N}=\Diff{M}{0}=0$ but $M0\neq0$ in general.
%This is crucial for the definition of the Taylor expansion.
We follow the tradition of quotienting also for the idempotency of $+$.
%Sums are, then, just \emph{finite} sets of terms.
Simple types are $A::= o\mid A\to A$.
The typing system is in Figure~\ref{fig:rules}, where a context $\Gamma$ is a list of typed variable declarations.
The axioms are $\Gamma, x:A \vdash x: A$ and $\Gamma\vdash 0:A$.

%In $\BSTLC$ the typing system handles duplications; in $\STDLC$ the syntax with its operational semantics (that we do not give) does it.
Writing $\Der^2[\_,(\_)^2]$ as a shortcut for $\Der[\Der[\_,\_],\_]$ and $\Der^1[\_,(\_)^1]$ for $\Diff{\_}{\_}$, the analogue of the previous $\BSTLC$-term is $\vdash_{\STDLC} \lambda {\color{red}z}. \Der^{\color{red}2}[
	\lambda x{\color{green}y}.
		\Der^{\color{violet}1} [
				\Der^{\color{blue}1} [y, x^{\color{blue}1}]
        0, x^{\color{violet}1}
	]0
, z^{\color{red}2}]0
: {\color{red}o}\to ({\color{green}o \to o \to o}) \to o$.
%Here we wrote $\Der^2[\_]\cdot (\_)^2$ as a shortcut for $\Der[\Der[\_]\cdot (\_)]\cdot (\_)$ and $\Der^1[\_]\cdot (\_)^1$ for $\Der[\_]\cdot (\_)$.
In particular, if the \emph{multiplicities} of the arguments (the colored exponents) do not exactly match the number of duplications, e.g.\ in $\vdash_{\STDLC} \lambda z. \Der^{\color{red}3}[
	\lambda xy.
		\Diff{
				\Diff{y}{x}
		0}{x}
	0
, z^{\color{red}3}]0
: o\to (o \to o \to o) \to o$, then the term reduces to the empty sum $0$ (representing an \emph{error}).

%Correspondingly, the syntax of the simply typed \emph{differential} $\lambda$-calculus ($\STDLC$) is defined by enriching $\STLC$ with a monoid structure $0,+$ over terms, as well as with $\Der$ and a notion of \emph{linear substitution} (see \cite{difflambda} or the Appendix for details).
%The models of $\STDLC$ are the cartesian \emph{closed} differential categories ($CC\partial C$), which are defined as $C\partial C$ which are also cartesian closed, and in which the monoid structure and the differential operator are both well-behaved with respect to the closed structure \cite{Manzo2012}.

The $\STDLC$ can be interpreted in Cartesian closed differential $\lambda$-categories (CC$\partial\lambda$C)\cite{Manzo2010,Blute2009, Blute2019}.
In them, homsets are equipped with a structure of commutative monoid %(i.e.\ we can add morphisms and there is a $0$ morphism)
%, it is Cartesian closed%(with the closed structure compatible with the additive structure)
and with a certain \emph{differential operator} $D$, generalising the usual notion of differentiation, see e.g.\ \cite{BluteEhrhTass10}.
%An example is the CC$\partial\lambda$C of convenient vector spaces with smooth maps, where $D$ is the ``real'' differential $Df:\mathbb{V}\times\mathbb{V}\rightarrow \mathbb{W}$, $Df(x,u):=\dfrac{d}{dt}{\!\Big|_{t=0}} f(x+tu)$, of smooth maps $f:\mathbb{V}\rightarrow \mathbb{W}$.
%More precisely, a cartesian category $\C C$ is a $C\partial C$ when:
%\begin{itemize}
%\item $\C C$ is left-additive, i.e.~its hom-sets have the structure of commutative monoids, and the cartesian structure is well-behaved w.r.t.~this monoid structure;
%\item $\C C$ is equipped with a differential operator $D:
%\C C(X,Y)\to \C C(X\times X,Y)$ satisfying some axioms which capture usual properties of differentials (e.g.~the linearity of $D$ in one of its two variables, the chain rule, etc.).
%\end{itemize}

Another intriguing similarity between program derivatives and 
actual derivatives is provided by the \emph{Taylor expansion} $\C T$:
%In analysis, a sufficiently regular function can be seen (locally around a point) as its Taylor expansion, i.e.\ a series of polynomials weighted via some rational coefficient.
%Now, a polynomial in $x$ of degree $k$ can be thought of as a function which uses its argument $x$ exactly $k$ times.
in $\STDLC$ (even untyped), we can perform a syntactic Taylor expansion of an ordinary $\lambda$-term via an inductively defined map $\Theta$ giving rise to an infinite series of terms, whose crucial case of the definition is
$
\Theta(MN)=\sum_{n\in\mathbb{N}}\frac{1}{n!}\Der^n\left[\Theta(M),\Theta(N)^n\right]0.
$
As in analysis, it decomposes an application as a series of $k$-linear functions, which can be seen as its approximants.
In other words, unbounded duplications correspond to some sort of limit of bounded, but arbitrarily large, ones.
Since we only consider idempotent sum, the factorial coefficient disappears, and the resulting Taylor map is called the \emph{qualitative} Taylor expansion, indicated by $\Te{\_}$.
% (this perspective is made clearer by the related approach of the \emph{resource $\lambda$-calculus} \cite{}).

@@

### 2023-07-21 CSL24/section3B.tex
Committ
Per ora è ancora molto copia ed incolla da LICS. Fai tutte le modifiche che vuoi, se vuoi. Io domani non posso lavorare, quindi prendi il lock quando vuoi. Da domenica, dimmelo così ci organizziamo. Notte !

@@

\begin{fact}
 $\LREL$ is Lafont.
\end{fact}

Let us recall that the coKleisli category $\C C_!$ of a category $\C C$ w.r.t.\ a comonad $!$ is the category whose elements are the same of $\C C$, and $\HOM{\C C_!}{X}{Y}:=\HOM{\C C}{!X}{Y}$, with composition $\circ_!$ defined by making use of the co-multiplication of $!$.
%We will explicit this constructions in our tropical setting in the next lines.
As already mentioned, one obtains a CCC from a Lafont category by defining the exponential objects as $X\to Y:=!X \multimap Y$.
In our case, specialising [Theorem III.7, \cite{Manzo2013}] we have indeed:
\begin{fact}
 The coKleisli $\LREL_!$ is CCC, i.e.\ a model of $\STLC$.
\end{fact}

%It is instructive what its CCC-structure looks like in our tropical world.
In our tropical setting, 
the exponential object $X\multimap Y$ is $!X\times Y$, and 
the coKleisli composition of $s\in\Lawv^{!Y\times Z}$ and $t\in\Lawv^{!X\times Y}$ is the matrix $s\circ_! t\in\Lawv^{!X\times Z}$ where $(s\circ_! t)_{\mu,c}$ is:
\begin{align}
 \inf\limits_{n\in\N, b_1\dots,b_n\in Y, \mu = \mu_1+\cdots +\mu_n}
 \left\{s_{[b_1,\dots,b_n],c} + \sum\limits_{i=1}^n t_{\mu_i,b_i}\right\}.
\end{align}
% (where $+$ is the disjoint union).

@@

As it is well-known, the Cartesian closed structure %of a category $\C C$ 
allows to define a sound interpretation $\model{\Gamma\vdash M:A}\in\HOM{\LREL_{!}}{\model \Gamma}{\model A}$ of terms as morphisms.
In our case, we have:

\begin{proposition}\label{prop:descrete}
 The interpretation $\model M$ of a $\lam$-term $M$ in $\LREL$ is a \emph{discrete} matrix, i.e.\ its coefficients are either $0$ or $\infty$.
\end{proposition}
\begin{proof}
 First easily prove that composition preserves discreteness.
 Then go by straightforward induction on $M$, using that the projections and evaluation of  $\LREL$ are discrete.
\end{proof}

While ordinary $\lam$-terms yield discrete matrices, 
as discussed in Section~\ref{sec:app}, the other coefficients from $[0,\infty]$ will appear when considering quantitative effects like computation steps or probabilities.
%, give rise to  we will however discuss how 
%we can use this model in order to make it keep track of quantitative effects of programs.

\begin{remark}\label{rmk:ModelsOfBSTLC}
%Similarly to It is known {\color{red}(([reference??] e dire meglio)} 
The exponential $!$ can be ``decomposed'' into a family of \emph{graded} exponentials $!_n:\LREL\to\LREL$ ($n\in\BB N$), defined by functors $X\mapsto \C M_{\leq n}(X)$. %  lift to functors 
The sequence $(!_n)_{n\in\N}$ can be then turned into a $\N$-graded linear exponential comonad on (the SMC) $\LREL$, satisfying the adjunction:
$\LREL(Z\otimes !_{n}X,Y) \simeq \LREL(Z, !_{n}X\multimap Y)$.
Therefore, $\LREL$, together with $(!_n)_{n\in\N}$, is a model for $\BSTLC$. 
In particular, bounded arrow types are interpreted via
%Concretely, the types of this calculus (Equation~\ref{eq:Btyped}) are interpreted as: 
$\model o=\{\star\}$ and 
$\model{!_{n}A\multimap B}:= \C M_{\leq n}(\model A) \times \model B$.
Notice that, for any type $A$ of $\BSTLC$, its interpretation $\model{A}$ is a finite set.
\end{remark}
%Bounded types are interesting because of the following proposition:
%\begin{proposition}
%For all bounded types $A,B$, the morphisms from $\model A$ to $\model B$ (in all parametric relational models) correspond to polynomials.
%\end{proposition}
%\begin{proof}
%It suffices to check that $\model A$ is finite for all bounded types $A$. Indeed this implies that a morphism $t:\model A\to \model B$ is a finite matrix $t: \model A \times \model B \to \Lawv$.Hence, its corresponding map $\widehat t:\Lawv^{\model A} \to \Lawv^{\model B}$ is a polynomial.
%\end{proof}

%For example (here $!_{n}(\Lawv^{X}):= \Lawv^{\C M_{\leq n}(X)}$):
%\begin{itemize}
%\item a map $f\in \LREL( !_{1}\Lawv, \Lawv)$ is of the form $f(x)=\min \{x+a,b\}$;
%\item a map $f\in \LREL(!_{2}\Lawv, \Lawv)$ is a ``quadratic'' polynomial $f(x)=\min\{2x+a, x+b, c\}$.
%\end{itemize}

>>>>

\begin{comment}

@@

\begin{proposition}
 $\LREL$ is Lafont.
\end{proposition}

\end{comment}

Let us recall that the coKleisli category $\C C_!$ of a category $\C C$ w.r.t.\ a comonad $!$ is the category whose elements are the same of $\C C$, and $\HOM{\C C_!}{X}{Y}:=\HOM{\C C}{!X}{Y}$, with composition $\circ_!$ defined by making use of the co-multiplication of $!$.
%We will explicit this constructions in our tropical setting in the next lines.
As well-known, if a SMCC is \emph{Lafont}, then one obtains a CCC from it by defining the exponential objects as $X\to Y:=!X \multimap Y$.
With no surprise $\LREL$ is Lafont (\cite[Corollary III.6]{Manzo2013}) and specialising \cite[Theorem III.7]{Manzo2013} we have indeed that the coKleisli $\LREL_!$ is CCC, i.e.\ a model of $\STLC$, where as usual $!$ acts on objects by taking the set of finite multisets.
%It is instructive what its CCC-structure looks like in our tropical world.
Remark that the cartesian product of $\LREL_!$ (and of $\LREL$) is the disjoint union $+$ of sets, the \emph{evaluation} morphisms are matrices in $\Lawv^{!(!X\times Y)+X)\times Y}$, and the coKleisli composition of $s\in\Lawv^{!Y\times Z}$ and $t\in\Lawv^{!X\times Y}$ is the matrix $s\circ_! t\in\Lawv^{!X\times Z}$ where $(s\circ_! t)_{\mu,c}:=
%\begin{align}
\inf_{n\in\N, b_1\dots,b_n\in Y, \mu = \mu_1+\cdots +\mu_n}
 \left\{s_{[b_1,\dots,b_n],c} + \sum_{i=1}^n t_{\mu_i,b_i}\right\}$.
%\end{align}
%where $+$ is the multiset union.

@@

%As it is well-known, the Cartesian closed structure %of a category $\C C$  allows to define a sound interpretation $\model{\Gamma\vdash M:A}\in\HOM{\LREL_{!}}{\model \Gamma}{\model A}$ of terms as morphisms.
%In our case, we have:
